---
title: "Seminar Paper"
author: "Jesse Keränen"
date: "5/7/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())
```

Load data and libraries and some constants
```{r}
library(data.table)
library(zoo)
library(ggplot2)
library(tidyr)
library(readxl)
library(tidyverse)
library(lubridate)
library(xtable)


load("/Users/jessekeranen/seminar/data/2022_GBR_DS_monthly.RData")
load("/Users/jessekeranen/seminar/data/2022_GBR_DS_static.RData")
load("/Users/jessekeranen/seminar/data/2022_GBR_WS_yearly.RData")

# Load data from Keneth French Website
temp <- tempfile()
base <- "https://mba.tuck.dartmouth.edu/pages/faculty/ken.french/ftp/"
factor <-"F-F_Research_Data_Factors"
format<- "_CSV.zip"
ff_url <- paste(base,factor,format,sep ="")
download.file(ff_url, temp)

FF_3_Factors_dt <- as.data.table(read_csv(unzip(temp),skip = 3) %>% 
rename(Date = ...1) %>%
  mutate_at(vars(-Date), as.numeric) %>%
  mutate(Date = ymd(parse_date_time(Date, "%Y%m")))) %>% 
  mutate(Date = lubridate::rollback(Date))
 
colnames(FF_3_Factors_dt) <- gsub("-", "_", colnames(FF_3_Factors_dt))

date_for_plot <- as.yearmon("Jul 2015")
```

```{r}
# Copy original data to a variable so that we don't need to change original data
dt <- copy(DS.monthly[, .(Id, ym, Date, RET.USD, MV.USD, RI.USD)])
dt[, month := month(Date)]
dt[, year := year(Date)]
dt[, hcol := ifelse(month >= 7, year-1, year-2)]
# Lagged market value
dt[, LMV.USD := shift(MV.USD, 1, type = "lag"), by = Id]

# Winsorize function, winsorizes given variables to 1st and 99th percentile
winsorize_border <- 0.01
winsorize <- function(dt, x, percentile){
  dt[, (x) := ifelse(get(x) > quantile(get(x), 1-percentile, na.rm=T), 
                     quantile(get(x), 1-percentile, na.rm=T), get(x)), by=Date]
  dt[, (x) := ifelse(get(x) < quantile(get(x), percentile, na.rm=T),
                     quantile(get(x), percentile, na.rm=T), get(x)), by=Date]
}
winsorize(dt, "RET.USD", winsorize_border)


# Merge yearly data to monthly data
dt <- merge(dt, WS.yearly[, .(Id, Year, WC05476, WC05001, WC01001, WC01051, 
                              WC01251, WC01101, WC05301, WC02999, WC03480, 
                              WC03451, WC03255, WC03251, WC01151, WC03051, 
                              WC03063, WC02201, WC02001, WC01551, WC03101,
                              WC03151, WC04551)], by.y = c("Id", "Year"),
            by.x = c("Id", "hcol"), all.x = T)

# For some variables we want to calculate yearly change. With help of this merge we get one year lagged values for these variables to same row as current
help2 <- dt[month == 7, .(hcol = hcol, Id = Id, LWC02999 = WC02999, 
                          LWC03151 = WC03151, LWC01151 = WC01151)]
help2[, hcol := hcol+1]
dt <- merge(dt, help2, by = c("Id", "hcol"), all.x = T)

# Book to market value
dt[, BM := (WC05476/WC05001)]
# Total book value = book value per share * shares outstanding
dt[, BE := WC05476*WC05301]
# Operating profitability
dt[, OP := WC01551/BE]
# Investment variable
dt[, INV := (WC02999 - LWC02999)/LWC02999]

# Order data.table by Date and decreasing by market value. Makes size allocation easier
setorder(dt, Date, -LMV.USD)

# Momentum variable
dt[, MOM := shift(RI.USD, 2, type = "lag")/shift(RI.USD, 12, type = "lag") - 1,
   by = Id]


size_breaks <- function(dt, quantiles, labels, name) {
  setorder(dt, Date, -LMV.USD)
  dt[, hcjun := ifelse(month >= 7, year, year-1)]
  
  temp <- dt[month==7 & !is.na(LMV.USD)]
  temp[, agg_weight := cumsum(LMV.USD)/sum(LMV.USD), by = year]
  temp[, tempr := cut(agg_weight, breaks = quantiles, labels = labels), Id]
  temp <- temp[, .(year, Id, tempr, agg_weight)]
  
  setnames(temp, "tempr", name)
  dt <- merge(dt, temp, by.x = c("Id", "hcjun"), by.y = c("Id", "year"), all.x = T)
}
dt <- size_breaks(dt, c(0, 0.9, 1), c("Big", "Small"), "pf.size")
panel_country <- copy(dt)


# Quantiles used in portfolio allocations
quantiles <- c(0, 0.3, 0.7, 1)

# Labels for momentum portfolios
labels <- c("Loser", "Neutral_mom", "Winner")
# We don't use factors function for momentum allocation because it differs a bit from other factors. Momentum allocation is done monthly where as other factors are constructed yearly basis.
panel_country[year > 1990, pf.mom := cut(MOM, breaks = quantile(.SD[pf.size == "Big", MOM], quantiles, 
                                                      na.rm = T), labels = labels, na.rm = T), by = Date]


# Function for allocating stocks to portfolios based on given variable
factor <- function(dt, variable, name, labels) {
  hlpvariable2 <- dt[month==7 & year > 1985, .(temp = cut(.SD[, get(variable)], breaks = 
                  quantile(.SD[pf.size =="Big", get(variable)], quantiles, na.rm = T), labels = 
                  labels), Id), by = year]
  
  setnames(hlpvariable2, "temp", name)
  # Merge the variable allocation back from July Y to June Y+1
  dt <- merge(dt, hlpvariable2, by.x=c("hcjun", "Id"), by.y=c("year", "Id"), all.x=T)
  
  return(dt)
}
panel_country <- factor(panel_country, "BM", "pf.bm", c("Low", "Neutral", "High"))
panel_country <- factor(panel_country, "OP", "pf.op", c("Weak", "Neutral_op", "Robust"))
panel_country <- factor(panel_country, "INV", "pf.inv", c("Conservative", "Neutral_inv", "Agressive"))

# Form the 2x3 portfolios combining size with one additional variable
panel_country[, Portfolio := paste0(pf.size,".",pf.bm)]
panel_country[, Portfolio2 := paste0(pf.size,".",pf.op)]
panel_country[, Portfolio3 := paste0(pf.size,".",pf.inv)]
panel_country[, Portfolio4 := paste0(pf.size,".",pf.mom)]


# Function for calculating portfolio returns
portfolioreturns <- function(dt, portfolio){
  portfolio_returns <- dt[, .(RET.USD = weighted.mean(RET.USD, LMV.USD, na.rm = T)),
                          by = c("Date", portfolio)]
  return(dcast(portfolio_returns, paste("... ~ ", portfolio)))
}
portfolio_returns1 <- portfolioreturns(panel_country, "Portfolio")
portfolio_returns2 <- portfolioreturns(panel_country, "Portfolio2")
portfolio_returns3 <- portfolioreturns(panel_country, "Portfolio3")


# For momentum we use equal weights
portfolio_returns4 <- panel_country[, .(RET.USD = mean(RET.USD, na.rm = T)),
                                    by = c("Date", "Portfolio4")]
portfolio_returns4 <- dcast(portfolio_returns4, ... ~ Portfolio4)


# Merge all portfolios to one
portfolio_returns <- merge(portfolio_returns1, portfolio_returns2, by = "Date")
portfolio_returns <- merge(portfolio_returns, portfolio_returns3, by = "Date")
portfolio_returns <- merge(portfolio_returns, portfolio_returns4, by = "Date")

portfolio_returns[is.na(portfolio_returns)] <- 0

# Market risk factor
market_return_dt <- dt[!is.na(RET.USD) & !is.na(LMV.USD), .(Return_Mkt = weighted.mean(RET.USD, LMV.USD)), by = "Date"]
market_return_dt <- merge(market_return_dt, FF_3_Factors_dt[, .(Date, RF)], by = "Date", all.x = T)
market_return_dt[, RMRF := Return_Mkt - RF]
RMRF_factor <- market_return_dt[, .(Date, RMRF = Return_Mkt-RF)]

# Calculate factors from portfolios
factors <- portfolio_returns[, .(Date, 
          SMBbm = (Small.High + Small.Neutral + Small.Low)/3 - (Big.High + Big.Neutral + Big.Low)/3,
          SMBop = (Small.Robust + Small.Neutral_op + Small.Weak)/3 - (Big.Robust + Big.Neutral_op +
                                                                        Big.Weak)/3,
          SMBinv = (Small.Conservative + Small.Neutral_inv + Small.Agressive)/3 - (Big.Conservative +
                                                                    Big.Neutral_inv + Big.Agressive)/3,
          SMBmom = (Small.Winner + Small.Neutral_mom + Small.Loser)/3 - (Big.Winner +
                                                                    Big.Neutral_mom + Big.Loser)/3,
          HML = (Small.High + Big.High)/2 - (Small.Low + Big.Low)/2,
          RMW = (Small.Robust + Big.Robust)/2 - (Small.Weak + Big.Weak)/2,
          CMA = (Small.Conservative + Big.Conservative)/2 - (Small.Agressive + Big.Agressive)/2,
          MOM = (Small.Winner + Big.Winner)/2 - (Small.Loser + Big.Loser)/2)
          ]

factors[, SMB := (SMBbm + SMBop + SMBinv + SMBmom)/4]

# Remove unnecessary columns
factors[, SMBop := NULL]
factors[, SMBbm := NULL]
factors[, SMBinv := NULL]
factors[, SMBmom := NULL]

# Merge market factor with other factors
factors_wide <- merge(factors, RMRF_factor, by="Date")

# Melt for plotting
factors <- melt(factors_wide, id.vars = "Date", variable.name = "factor", value.name = "RET.USD")

# Check the factor statistics
factors[factor == "SMB",t.test(RET.USD)]
factors[factor == "HML",t.test(RET.USD)]
factors[factor == "RMW",t.test(RET.USD)]
factors[factor == "CMA",t.test(RET.USD)]
factors[factor == "MOM",t.test(RET.USD)]
factors[factor == "RMRF",t.test(RET.USD)]

# Just for plotting
factors2 <- factors[Date >= "1990-07-31", .(Date, cum_prod = cumprod(RET.USD/100 + 1)), by = factor]
ggplot(factors2, aes(Date, cum_prod)) + geom_line(aes(color = factor)) +
  labs(y = "Cumulative return", title = "") + scale_y_log10()
```


Size allocation:
```{r}
dt2 <- dt[, .(Id, ym, Date, RET.USD, MV.USD, month, year, LMV.USD, BM, OP, INV, MOM, pf.size)]

quantiles <- c(0, 0.75, 0.90, 0.96, 0.99, 1)
dt2 <- size_breaks(dt2, quantiles, c(5, 4, 3, 2, 1), "portf.size")

# Plot to see if our allocation seems reasonable
options(scipen = 999)

# Cumulutave distribution of the market values in 2015 allocation
ggplot(dt2[ym == date_for_plot], aes(LMV.USD, agg_weight)) + geom_line() + scale_x_log10()
# Market values of stocks in different portfolios don't seem to overlap
ggplot(dt2[ym == date_for_plot], aes(as.factor(portf.size), LMV.USD)) + geom_boxplot() + scale_y_continuous(trans='log10')
```

Allocation based no Book to Market-ratio. I tried to look at WC05001 to see if it consistent with stock prices from internet for some stocks. For me it seems like these values are somehow scaled.
```{r}
quantiles <- seq(0, 1, by = 1/5)
dt2 <- factor(dt2, "BM", "portf.bm", c(1, 2, 3, 4, 5)) 

ggplot(dt2[ym == date_for_plot], aes(as.factor(portf.bm), BM)) + geom_boxplot() + scale_y_log10()
```

By combining Size and Be/Me allocations, each stock will be allocated one of the 25 portfolios.
```{r}
dt2[, portfolio := paste(as.character(portf.size), as.character(portf.bm), sep = " ")]

# Remove observation that weren't allocted to a portfolio
dt2 <- dt2[!grepl("NA", portfolio)]

# Let's see if there is reasonable amount of stocks in each portfolio each year. Now we can't expect there to be equal amount of stocks in each portfolio if size and value correlate. Also we didn't use equal breakpoints for size.
tail(table(dt2[year(ym) == 2015, portfolio, by = ym]))
```

Portfolio returns
```{r}
portfolios <- portfolioreturns(dt2, "portfolio")

# Because unbalanced portfolio construction there for some months there can be zero stocks in some portfolio
portfolios[is.na(portfolios)] <- 0


#For plotting
portfolios2 <- portfolios[year(Date) > 1990]

portfolios2 <- melt(portfolios2, id.vars = "Date", variable.name = "Portfolio",
     value.name = "Return")
portfolios2[, RET.CUM := cumprod(Return/100 + 1), by = Portfolio]
portfolios2 <- merge(portfolios2, market_return_dt[, .(Date, RF)], by = "Date")
portfolios2[, EXCESS.RET := Return - RF]

ggplot(portfolios2[year(Date) > 1990], aes(Date, RET.CUM, color = Portfolio)) + geom_line() + scale_y_log10()
```

Data seems weird. For example company id 882323, from 2011-12-31 to 2012-10-31.

```{r, results='asis'}
p <- portfolios2[, mean(Return), by = Portfolio]
separate(p, "Portfolio", into = c("Size", "Value"), sep = " ") %>% spread(Value, V1)
```

```{r, results='asis'}
library(stargazer)

portfolio_names <- portfolios2[, unique(Portfolio)]
# One can try different models by choosing different subset of factors here
fac <- as.matrix(factors_wide[year(Date) > 1990, -1])

list <- vector("list", length(portfolio_names))

for (i in 1:length(portfolio_names)){
  list[[i]] <- lm (portfolios2[Portfolio == portfolio_names[i], EXCESS.RET] ~ fac)
}

stargazer(list[1:8], type='latex', header=FALSE, no.space = TRUE,
           column.sep.width = "1pt", font.size = "tiny")

stargazer(list[9:16], type='latex', header=FALSE, no.space = TRUE,
           column.sep.width = "1pt", font.size = "tiny")

stargazer(list[17:25], type='latex', header=FALSE, no.space = TRUE,
           column.sep.width = "1pt", font.size = "tiny")
```

